Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-12 manuscript_rmarkdown/manuscript.tex

Location- and scale-invariant transformations are realised by
simultaneously optimising three parameters, i.e.~transformation
parameter \(\lambda\), shift parameter \(x_0\) and scale parameter
\(s\). We derived the log-likelihood function to facilitate optimisation
using MLE. Alternatively, standardisation of a numeric variable (e.g.,
through subtracting its median value and division by its interquartile
range) prior to conventional power transformations can achieve a similar
effect in reducing sensitivity to the feature's location and scale.
While this alternative helps prevent these issues -- provided that
standardisation does not lead to negative values for Box-Cox
transformations -- location- and scale-invariant transformations are
able to achieve better transformations to normality, as demonstrated by
lower residual errors in simulation experiments.

We assessed several weighting methods to achieve robust power
transformations. Methods that relied on the z-score of the transformed
feature or the residual error yielded worse results than the non-robust
method. Underperformance of these weighting methods could be explained
by their reliance on transformed feature values for setting weights.
Consequently, their weights change at each iteration in the MLE
optimisation process. This increases local variance in the
log-likelihood function and creates local optima that the optimiser may
not handle well. Methods that relied on the empirical probability did
not suffer from this issue, as weights remained fixed during MLE.

>>>>

Location- and scale-invariant transformations are realised by
simultaneously optimising three parameters, i.e.~transformation
parameter \(\lambda\), shift parameter \(x_0\) and scale parameter
\(s\). We derived the log-likelihood function to facilitate optimisation
using MLE. Alternatively, standardisation of a numeric feature (e.g.,
through subtracting its median value and division by its interquartile
range) prior to conventional power transformations can achieve a similar
effect in reducing sensitivity to the feature's location and scale.
While this alternative helps prevent these issues -- provided that
standardisation does not lead to negative values for Box-Cox
transformations -- location- and scale-invariant transformations are
able to achieve better transformations to normality, as demonstrated by
lower residual errors in simulation experiments.

We assessed several weighting methods to achieve robust power
transformations. Robust power transformations should satisfy two
conflicting aims: they should minimise residual errors after
transformation to central normality, and minimise the overall effect of
outliers on estimation of transformation parameter \(\lambda\). These
aims are reflected in the composite loss used to optimise weighting
parameters. For Yeo-Johnson transformation, empirical probabilities with
tapered cosine weighting resulted in the lowest loss. The optimal
parameters led to weights that gradually decline towards the tails of
empirical probabilities, i.e.~elements with low and high values receive
less or no weight. Similarly, for Box-Cox transformation, empirical
probabilities with step weighting resulted in the lowest loss. In this
case the central 80\% of the elements are used, and the remaining 10\%
in each tail are ignored. Methods that relied on the z-score of the
transformed feature or the residual error yielded worse loss than the
non-robust method or those based on empirical probabilities.
Underperformance of these weighting methods could be explained by their
reliance on transformed feature values for setting weights.
Consequently, their weights change at each iteration in the MLE
optimisation process. This increases local variance in the
log-likelihood function and creates local optima that the optimiser may
not handle well. Methods that relied on the empirical probability did
not suffer from this issue, as weights remained fixed during MLE.
